---
title: "STAT/MATH 495: Problem Set 09"
author: "Brendan Seto"
date: "2017-11-07"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    collapsed: false
    smooth_scroll: false
    df_print: kable
    code_fold: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.width=8, fig.height=4.5, message=FALSE, warning = FALSE
  )
set.seed(76)

library(tidyverse)
library(mosaic)
library(gridExtra)
```


# Question 1: Run k-means

```{r}
observations_1 <- read_csv("data/observations_1.csv")
observations_2 <- read_csv("data/observations_2.csv")

# Set observations to be one of two datasets
observations <- observations_2

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

p2 <-ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5) +
  labs(list(title = "Data from Observations_2"))+
  guides(col=FALSE)
```

```{r}
# Set observations to be one of two datasets
observations <- observations_1

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

p1 <- ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5) +
  labs(list(title = "Data from Observations_1"))+
  guides(col=FALSE)

grid.arrange(p1,p2, ncol=2)
```

##Answers:

###1. Run KMC 10 times on `observations_1` and comment on the consistency of the results.

For the purposes of this assignment, I will only be looking at the location of the cluster centers (i hope this is right).

```{r}
kms1 <- function(data){
  # Set observations to be one of two datasets
  observations <- data

  # Fit model for k=2
  k <- 2
  k_means_results <- kmeans(observations, centers=k)
  clusters <- k_means_results$cluster
  cluster_centers <- k_means_results$centers

  # Add cluster results to observations. Note we convert to factor since cluster
  # ID's should be treated as categorical
  observations$cluster <- as.factor(clusters)

  # Add cluster ID's to cluster_centers
  cluster_centers <- cluster_centers %>% 
    as_tibble() %>% 
    arrange(x1,x2) %>% 
    mutate(cluster=as.factor(1:k))
  
  return(cluster_centers)
}

prob1 <- do(10)*kms1(observations_1)

prob1 %>% group_by(cluster) %>% 
  summarise(mean_x1 = mean(x1), mean_x2 = mean(x2), sd_1 = sd(x1), sd_x2 = sd(x2))
```

The results are not very consistent.  Were I to run a t.test, I would find that there is no significant difference between clusters.  

###2. Speculate on the root cause of any consistency or inconsistency in the results.

I don't think there is a "natural" grouping in this data.  As we can see in the plots above, there is a fairly uniform distribution.  I do not expect distinct clusters to come up every time.

###3. Run KMC 10 times on `observations_2` and comment on the consistentcy of the results.

```{r}
prob2 <- do(10)*kms1(observations_2)

prob2 %>% group_by(cluster) %>% 
  summarise(mean_x1 = mean(x1), mean_x2 = mean(x2), sd_1 = sd(x1), sd_x2 = sd(x2))

```

Here there is no standard deviation, the same clusters appear every time.  This leads me to believe that there is a very distinct pair of clusters that are so strong, they appear over and over again.  

###4. Speculate on the root cause of any consistency or inconsistency in the results.

Look at the graph.  There are clearly two distinct groupings.  I bet Albert simulated the data from these two points, but even then it's striking that they are exactly the same every time...

# Bonus question: Code your own

Read ISLR page 388 Algorithm 10.1 and implement k-means clustering from scratch.
Don't worry about doing it for general $k$; keep it simple and do it for $k=2$
specifically. Apply it to `observations_2` from above.

```{r, echo = TRUE}
# Hint:
library(proxy)

A <- data_frame(
  x1 = c(0, 0.5, 0.75, 1),
  x2 = c(0, 0.5, 0.75, 1)
  )

B <- data_frame(
  x1 = c(1, 0),
  x2 = c(1, 0)
  )

kms2_pre <- function(data, center1, center2){
  A <- data
  B <- data_frame(
    x1 = center1,
    x2 = center2
  )
  distance_matrix <- proxy::dist(x=A, y=B)
  distance_matrix
  data <- data %>% mutate(cluster = apply(distance_matrix, 1, which.min)) %>% 
    group_by(cluster) %>% 
    summarise(x1 = mean(x1), x2 = mean(x2))
  
  return(data)
}

for(i in seq(1:10)){
  B <- kms2_pre(observations_2, B$x1, B$x2) %>% select(x1,x2)
}

B %>% arrange(x1) %>% cbind(prob2 %>% group_by(cluster) %>% summarise(P2_x1 = mean(x1), P2_x2 = mean(x2)))
```

Yay! Got same clusters!